Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-14 talk/iwtc11/paper.tex

The original trace (prior too peeling) consists of three parts. 
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the pereamble, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

In the general case, each virtual in the jump arguments is exploded into a
vector of variables containing the values of all registered attributes. If some
of the attributes are themselves virtuals they are recursively exploded
to make the vector contain only non virtual variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every non
virtual variable of the jump arguments is exploded into a vector. This will
be a vector containing the original variable only. To summarize, for
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
        H^{\left(k\right)} & \text{if $J_k$ is virtual}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all non virtual
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
and the arguments of the \lstinline{jump} operation of the peeled loop,
$K$, constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

The loop peeling optimization was implemented in the PyPy
framework. That means that the jit compilers generated for all
interpreters implemented within PyPy now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
would be short numerical calculations with no failing guards and no
external calls.

\subsection{Python}
The python interpreter of the PyPy framework is a complete python
version 2.7 compatible interpreter. A set of numerical
calculations where implemented in both python and in C and their
runtimes compared. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The later, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one dimensional convolution with a kernel of fixed
  size $3$.
\item {\bf conv5}: one dimensional convolution with a kernel of fixed
  size $5$.
\item {\bf conv3x3}: two dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two dimensional
  arrays.
\item {\bf dilate3x3}: two dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: an low level video processing algorithm used to
  locate edges in an image. It calculated the gradient magnitude
  using sobel derivatives. The algorithm is in python implemented
  on top of a custom image class that is specially designed for the
  problem. It ensures that there will be no failing guards, and makes
  a lot of the two dimension index calculations loop invariant. The
  intention here is twofold. It shows that the performance impact of
  having wrapper classes giving objects some application specific
  properties can be negligible. This is due to the inlining performed
  during the tracing and the allocation removal of the index objects
  introduced. It also shows that it is possible to do some low level
  hand optimizations of the python code and hide those optimization
  under a nice interface without loosing performance.
\end{itemize}

>>>>

The original trace (prior to peeling) consists of three parts.
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the pereamble, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

In the general case, each virtual in the jump arguments is exploded into a
vector of variables containing the values of all registered attributes. If some
of the attributes are themselves virtuals they are recursively exploded
to make the vector contain only non-virtual variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every non-
virtual variable of the jump arguments is exploded into a vector. This will
be a vector containing the original variable only. To summarize, for
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
        H^{\left(k\right)} & \text{if $J_k$ is virtual}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all non virtual
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
and the arguments of the \lstinline{jump} operation of the peeled loop,
$K$, constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

The loop peeling optimization was implemented in the PyPy
framework. That means that the JIT-compilers generated for all
interpreters implemented within PyPy now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
would be short numerical calculations with no failing guards and no
external calls.

XXX reason why we use small numerical kernels for benchmarks

XXX we either need to explain that we use C++ or consistently use C

\subsection{Python}
The python interpreter of the PyPy framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python and in C and their
runtimes compared. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The latter, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In Python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one-dimensional convolution with a kernel of fixed
  size $3$.
\item {\bf conv5}: one-dimensional convolution with a kernel of fixed
  size $5$.
\item {\bf conv3x3}: two-dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two-dimensional
  arrays.
\item {\bf dilate3x3}: two-dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: a low-level video processing algorithm used to
  locate edges in an image. It calculates the gradient magnitude
  using sobel derivatives. In Python the algorithm is implemented
  on top of a custom image class that is specially designed for the
  problem. It ensures that there will be no failing guards, and makes
  a lot of the two dimension index calculations loop invariant. The
  intention here is twofold. It shows that the performance-impact of
  having wrapper classes giving objects some application-specific
  properties can be negligible. This is due to the inlining performed
  during the tracing and the allocation removal of the index objects
  introduced. It also shows that it is possible to do some low-level
  hand optimizations of the Python code and hide those optimization
  under a nice interface without loosing performance.
\end{itemize}

XXX we need Psyco numbers
